---
title: "Polynomial regression model: exercices"
---


## Preliminary

Only functions from `R`-base and stats (preloaded) are required plus packages from the **tidyverse** for data representation and manipulation. You could also try the package **broom** that standardizes the output of built-in R functions for statistical modelling


```{r linear-regression-config, message = FALSE}
library(tidyverse)
library(ggfortify) # extend some ggplot2 features
library(broom)
theme_set(theme_bw())
```

## Introduction

The file `ratWeight.csv` consists of rat weights measured over 14 weeks during a subchronic toxicity study related to the question of genetically modified (GM) corn. 

We will only consider the weight of rat B38625. 

```{r linear-regression-read, message = FALSE}
rat_weight <- read_csv("ratWeight.csv") %>% filter(id == 'B38625')
rat_weight %>% kableExtra::kbl() %>% kableExtra::kable_styling()
```

Based on this data, our objective is to build a regression model of the form

$$y_j = f(x_j) + e_j \quad ; \quad 1 \leq j \leq n$$

We will restrict ourselves to polynomial regression, by considering functions of the form 

$$
\begin{aligned}
f(x) &= f(x ; c_0, c_1, c_2, \ldots, c_d) \\
&= c_0 + c_1 x + c_2 x^2 + \ldots + c_d x^d
\end{aligned}
$$

## Questions

  1. Plot the data

  2. Fit several polynomials (degree 0, 1, 2, 3, 6, ...) to this data using the `lm` function. 

  3. Look at some diagnostic plots to eliminate miss-specified models.

  4. Compare the predictive performance of the models that have been retained.

  5. Use statistical tests and information criteria to compare the models and select *"the best one"*.

