---
title: 'Linear Mixed Models'
subtitle: 'Exercices'
---

## Preliminary {.unnumbered}

Functions from `R`-base and stats (preloaded) are required plus packages from the **tidyverse** for data representation and manipulation. We also need the **lme4** package which is the standard for fitting mixed-model. **lattice** is used for graphical representation of quantities such as random and fixed effects in the mixed models.

```{r tests-config, message = FALSE}
#| code-fold: false
library(tidyverse)
library(lme4)
library(lattice)
theme_set(theme_bw())
```


## The `Orange` data

The `Orange` data is a classical data set provided with the **datasets** package included with `R`. It consists in a 35-row data frame with  records of the growth of 3 orange trees, giving their circumference against their age: 


```{r orange-data-loading}
data(Orange)
## we force the sorting of the levels (1,2,3)
Orange$Tree <- factor(Orange$Tree, levels = unique(Orange$Tree))
Orange %>% as_tibble() %>% rmarkdown::paged_table()
```

```{r, fig.height=6, fig.width=8}
#| code-fold: true
pl <- 
  Orange %>% 
    ggplot() + aes(x = age, y = circumference, color = Tree) + 
  geom_point() + geom_line() 
pl
```

1. Fit different linear (fixed effect) models to this data and compare them
```{r}
lm1 <- lm(circumference~age+0, data=Orange) 
pl + geom_line(aes(x = age, y = predict(lm1))) + facet_wrap(~Tree) + theme_bw()
```
```{r}
lm2 <- lm(circumference~age, data=Orange) 
pl + geom_line(aes(x = age, y = predict(lm2))) + facet_wrap(~Tree) + theme_bw()
```
```{r}
# compare these two models 
anova(lm1, lm2)
BIC(lm1, lm2)
AIC(lm1, lm2)
# it seems that lm1 is better 
```

2. Fit different linear mixed effect models to this data and compare them (with/without random effect for the slope/intercept)
```{r}
# random effect only on intercept
lmem1 <- lmer(circumference ~ age + (1|Tree), data = Orange) 
# random effect only on slope
lmem2 <- lmer(circumference ~ age + (age + 0|Tree), data = Orange)
# random effect on intercept and slope
lmem3 <- lmer(circumference ~ 1+age + (1+age|Tree), data = Orange)
```
```{r}
BIC(lmem1, lmem2, lmem3, lm1, lm2)
# it seems that lmem2 (random effect only on slope) is the best.
# lmem2 and lmem3 do the similar prediction. BIC choose the one with less parameters. 

# Attention to the df! 
# lm1: sigma(variance of noise), intercept
# lm2: sigma(variance of noise), intercept, slope of age
# lmem1: sigma(variance of noise), intercept, slope of age, variance of random effects on Tree for intercept 
# lmem3: sigma(variance of noise), intercept, slope of age, 
        # variance of random effects on Tree for slope of age, 
        # variance of random effects on intercept,
        # covariance of these two above
```

```{r}
pl + geom_line(aes(x = age, y = predict(lmem2))) + facet_wrap(~Tree) + theme_bw()
```

3. Compute confidence intervals on the parameters, compute the individual parameters and plot the random effects for the *best* model
```{r}
confint(lmem2)
```

4. Compare the ML and the REML estimates for this model
```{r}

```

## The `Pastes` data

The `Pastes` data is a another classical data set provided with the **lme4** package. It describes the 
strength of a chemical paste product whose quality depends on the delivery batch, and the cask within the delivery. There are 60 observations, 10 levels of batch and 3 levels of cask.

```{r pastes data loading}
data(Pastes)
Pastes %>% as_tibble() %>% rmarkdown::paged_table()
```
We can have a quick look on summary statistics in each level of each factor:

```{r tidy alternative cross}
Pastes %>% 
  group_by(batch, cask) %>% 
  summarise(mean_strengh = mean(strength), sd_strength = sd(strength), n_replicat = n()) %>% 
  rmarkdown::paged_table()
```

There is obviously some structure in the data set and an effect of the batch level:

```{r Paste plot}
#| code-fold: true
Pastes %>% 
  ggplot() + aes(x = fct_reorder(batch, strength, median), y = strength) + geom_boxplot() + geom_jitter()
```

We can plot the cask as the function of the strength, conditionning on batch (and )
the later being reorder by strength)

```{r Pastes plot 2}
#| code-fold: true
lattice::dotplot(cask ~ strength | reorder(batch, strength), Pastes,
        strip = FALSE, strip.left = TRUE, layout = c(1, 10),
        ylab = "Cask within batch",
        xlab = "Paste strength", jitter.y = TRUE)
```

1. Build simple linear models for this data (with batches, then with casks effects)
```{r}
lm0 <- lm(strength~1, data = Pastes)
lm1 <- lm(strength~batch + 0, data = Pastes)
lm2 <- lm(strength~batch + batch:cask + 0, data = Pastes)
```

```{r}
BIC(lm0, lm1, lm2)
```

2. Build mixed models for this data (with random effects on batches, then casks effects)
```{r}
lmem1 <- lmer(strength ~ 1 + (1|batch), data = Pastes) 
lmem2 <- lmer(strength ~ 1 + (1|batch:cask), data = Pastes) 
```

3. What is the main cause of the variability of the paste strength?

## The `Oats` data

These data have been introduced by Yates (1935) as an example of a split-plot design. The objective is to assess the effect of treatment (level of nitrogen) on the yield of different varieties of oat.

The treatment structure used in the experiment was a 3 x 4 full factorial, with three varieties of oats and four concentrations of nitrogen. The experimental units were arranged into six blocks, each with three whole-plots subdivided into four subplots. The varieties of oats were assigned randomly to the whole-plots and the concentrations of nitrogen to the subplots. All four concentrations of nitrogen were used on each whole-plot.

The data are available from the **nlme** package:

```{r, message=FALSE, warning=FALSE}
data(Oats, package = "nlme")
# keep the original order of the levels of Block and Variety
Oats$Block   <- factor(Oats$Block  , levels = unique(Oats$Block))  
Oats$Variety <- factor(Oats$Variety, levels = unique(Oats$Variety))
Oats %>% as_tibble() %>% rmarkdown::paged_table()
```
The `xtabs` function creates a contingency table using the handy formula notation which allows us to check that the data indeed follows a full factorial design:

```{r}
xtabs(~ Block + Variety, Oats)
```

1. Plot the data in such a way as to visualize the effect of the fertilizer-concentration on the yield as well as possible differences between blocks and varieties.

2. Fit the Oats data with linear mixed effects model, assuming a linear effect of the fertilizer concentration on the yield.
