---
title: "Mixture models: exercices"
---

## Preliminary

Only functions from `R`-base and stats (preloaded) are required plus packages from the **tidyverse** for data representation and manipulation.

```{r mixture-model-lab-config, message = FALSE}
library(tidyverse)
theme_set(theme_bw())
```

## Faithful data

The `faithful` data consist of the waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.


```{r mixture-faitful-load}
data("faithful")
faithful %>% 
  ggplot() + aes(x=waiting) + geom_density() + xlab("waiting (mm)")
```

Fit variants of a mixture of two Gaussian distributions to the faithful data.

 1. assuming different proportions, means and variances for the 2 distributions

 2. assuming same variances 
 
 3. assuming same means
 
 4. assuming same proportions

### Hint: skeleton of EM for 1D mixture of Gaussian

We follow what has been done during the course: we define the function that computes the value of the likelihood for each point in each component. It will be used to compute the posterior probabilities in the E-step.

```{r auxiliary function}
## likelihood (density) of each point in each component
## data (x) is fixed while theta is variable
dcomponents <- function(theta, x) {
    mapply(
      function(pi, mu, sigma) pi * dnorm(x, mu, sigma),
      theta$pi, theta$mu, theta$sigma,
      SIMPLIFY = TRUE
    )
}
```

Then, we define an EM where the function that computes the M step is given as an argument:

```{r general EM}
mixture_gaussian1D <- 
  function(x,      # a n-vector of data
           theta0, # starting values of vector parameter
           M_step    = M_step_general, # an R function
           max_iter  = 100,
           threshold = 1e-6) {

  ## initialization
  n <- length(x)
  likelihoods  <- dcomponents(theta0, x) 
  deviance     <- numeric(max_iter)
  deviance[1]  <- -2 * sum(log(rowSums(likelihoods)))

  for (t in 1:max_iter) {
    
    # E step
    tau <- likelihoods / rowSums(likelihoods)

    # M step
    theta <- M_step(tau, x)
    
    ## Assessing convergence
    likelihoods   <- dcomponents(theta, x)
    deviance[t+1] <- - 2 * sum(log(rowSums(likelihoods)))

    ## prepare next iterations
    if (abs(deviance[t + 1] - deviance[t]) < threshold)
      break
    
  }

  list(theta = theta, deviance = deviance[t + 1])
}
```

### 1. assuming different proportions, means and variances for the 2 distributions
```{r}
# for each variant, M_step_general is different.
# (The M-step is different) because we add some hypothesis.
M_step_general_1 <- function(tau, x){
  pi <- colMeans(tau)
  mu <- colSums(tau*x) / colSums(tau)
  sigma <- sqrt(colSums(tau*x^2) / colSums(tau) - mu^2)
  list(pi = pi, mu = mu, sigma = sigma)
}

out <- mixture_gaussian1D(faithful$waiting, theta0=list(pi=c(.5,.5), mu=c(50,80), sigma=c(5,5)),M_step= M_step_general_1)
out$theta
out$deviance
```

```{R}
# BIC
out$deviance + log(length(faithful$waiting)) * 5
# 5 parameters : 1 pi, 2 means mu, 2 variance sigma
```

### 2. assuming same variances
```{R}
M_step_general_2 <- function(tau, x){
  pi <- colMeans(tau)
  mu <- colSums(tau*x) / colSums(tau)
  sigma <- sqrt(mean(x^2) - sum(pi * mu^2))
  list(pi = pi, mu = mu, sigma = sigma)
}
out_2 <- mixture_gaussian1D(faithful$waiting, theta0=list(pi=c(.4,.6), mu=c(50,80), sigma=c(5,5)), M_step = M_step_general_2)

out_2$theta
out_2$deviance
```
```{r}
out_2$deviance + log(length(faithful$waiting)) * 4
# 4 parameters : 1 pi, 2 means mu, 1 variance sigma
```

### 4. assuming same proportions
```{r}
M_step_general_4 <- function(tau, x){
  pi <- c(.5,.5)
  mu <- colSums(tau*x) / colSums(tau)
  sigma <- rep(sqrt(mean(x^2) - sum(pi * mu^2)), 2)
  list(pi = pi, mu = mu, sigma = sigma)
}
out_4 <- mixture_gaussian1D(faithful$waiting, theta0=list(pi=c(.5,.5), mu=c(50,80), sigma=c(5,5)),M_step = M_step_general_4)
out_4$theta
out_4$deviance
```
```{r}
out_4$deviance + log(length(faithful$waiting)) * 4
# 4 parameters : 2 means mu, 2 variance sigma
```

## Epilepsy data

The data `seizures.csv` consists of  daily counts of epileptic seizures for 6 patients. 

```{r mixture-seizures-load, message = FALSE}
seizures <- readr::read_csv('seizures.csv')
seizures %>% 
  ggplot() + aes(x = time, y = nsz) + geom_point() + facet_wrap( ~ id)
```
 
  1. Select the id 12 and fit a Poisson distribution to the number of seizures for this patient.

  2. Implement and use a EM algorithm for fitting a mixture of 2 Poisson distributions to this data
  
  3. Compare the two models
  
