---
title: "MAP566 - Homework assignment #1"
format:
  html:
    self-contained: true
    theme: [cosmo, theme.scss]
    toc: true
    number-sections: true
    html-math-method: katex
    code-copy: true
    code-summary: "Show the code"
    code-overflow: wrap
---

```{r, echo = FALSE, message=FALSE, results='hide', warning=FALSE}
library(tidyverse)
library(pwr) # power study 
library(saemix)
```

# Hypothesis testing: Toxicity assessment of the MON810 maize

## Introduction

The data set `MON810.csv` consists of several measurements made during a subchronic toxicity study concerning the MON810 maize.

```{r}
MON810 <- readr::read_csv("MON810.csv", show_col_types = FALSE)
rmarkdown::paged_table(MON810)
```

Biochemical parameters reflecting most physiological functions were measured two times (week 5 and 14), in particular through serum and urine chemistry, and hematology. Organ weights were measured at week 14.

The main objective of this study is to evaluate possible GMO effects on these parameters.

## Single comparison

We consider the variable "CALCIUM".
    
  1. Test if the mean level of calcium for period 2 is the same for males and females (*hint:* plot first the data and justify the test(s) to use).
  

```{r}
# plot the calcium level of period 2 for males and females

MON810_clean <- MON810 %>% filter (!is.na(`CALCIUM blood`))

MON810_clean %>% filter (period == 2) %>% 
  ggplot(aes(x=sex, y=`CALCIUM blood`, colour=`CALCIUM blood`, fill=`CALCIUM blood`)) + 
  geom_boxplot(alpha=0.5, outlier.alpha=0) +
	geom_jitter(width=0.25) +
	stat_summary(fun=mean, colour="red", geom="point",shape=18, size=3) + #red point is the mean
  xlab("sex") + 
  ylab("CALCIUM blood") +
  ggtitle('level of calcium in period 2 for males and females') +
  theme_bw()
```
   
### Question 1
Assume normal distribution for CALCIUM level in blood, we can perform student test.
```{r}
alpha <- 0.05
x <- MON810_clean %>% filter (period == 2 & sex == 'F') %>% pull("CALCIUM blood")
y <- MON810_clean %>% filter (period == 2 & sex == 'M') %>% pull("CALCIUM blood")
cat('Variance of x is ', var(x), '\n') 
cat('Variance of y is ', var(y), '\n') 
# we assume different variance, based on the variance of samples
t.test(x, y, conf.level = 1-alpha, var.equal = FALSE)
```
The p-value is quite small. We reject the null hypothesis. At level alpha=0.05, we could not prove that the mean level of calcium for period 2 is the same for males and females. It could also be observed in the figure above.


    
    
  2. test for the males if the mean level of calcium is the same for period 1 and period 2.
```{r}
MON810_clean %>% filter (sex == 'M') %>% 
  ggplot(aes(x=as.factor(period), y=`CALCIUM blood`, colour=`CALCIUM blood`, fill=`CALCIUM blood`)) + 
  geom_boxplot(alpha=0.5, outlier.alpha=0) +
	geom_jitter(width=0.25) +
	stat_summary(fun=mean, colour="red", geom="point",shape=18, size=3) + #red point is the mean
  xlab("period") + 
  ylab("CALCIUM blood") +
  ggtitle('level of calcium for males in period 1 and period 2') +
  theme_bw()
```
```{r}
alpha <- 0.05
x <- MON810_clean %>% filter (period == 1 & sex == 'M') %>% pull("CALCIUM blood")
y <- MON810_clean %>% filter (period == 2 & sex == 'M') %>% pull("CALCIUM blood")
cat('Variance of x is ', var(x), '\n') 
cat('Variance of y is ', var(y), '\n') 
# we assume different variance, based on the variance of samples
t.test(x, y, conf.level = 1-alpha, var.equal = FALSE)
```

### Question 2
We can reject the null hypothesis. At level alpha=0.05, we could not prove that for the males if the mean level of calcium is the same for period 1 and period 2.


  3. test for the males if the mean level of calcium for period 2 is the same for the control group and the MON810 group.
```{r}
MON810_clean %>% filter (sex == 'M' & period == 2 & (regimen == 'control' | regimen == 'MON810')) %>% 
  ggplot(aes(x=regimen, y=`CALCIUM blood`, colour=`CALCIUM blood`, fill=`CALCIUM blood`)) + 
  geom_boxplot(alpha=0.5, outlier.alpha=0) +
	geom_jitter(width=0.25) +
	stat_summary(fun=mean, colour="red", geom="point",shape=18, size=3) + #red point is the mean
  xlab("regimen") + 
  ylab("CALCIUM blood") +
  ggtitle('level of calcium for males in period 2, for control group and MON810 group') +
  theme_bw()
```
```{r}
alpha <- 0.05
x <- MON810_clean %>% filter (period == 2 & sex == 'M' & regimen == 'control') %>% pull("CALCIUM blood")
y <- MON810_clean %>% filter (period == 2 & sex == 'M' & regimen == 'MON810') %>% pull("CALCIUM blood")
cat('Variance of x is ', var(x), '\n') 
cat('Variance of y is ', var(y), '\n') 
# we assume different variance, based on the variance of samples
t.test(x, y, conf.level = 1-alpha, var.equal = FALSE)
```

### Question 3
The p-value is quite big. At level alpha=0.05, we could say that the mean level of calcium in period 2 for the males is the same for the control group and the MON810 group.


  4. What is the probability to detect a difference of one standard deviation (1 sd) with only 10 animals per group? with 20 animal? How can we ensure to detect such difference with a probability of 80%?
```{r}
alpha <- 0.05
pwr.t.test(n = 10, d = 1, type = "two.sample",
           alternative="two.sided", sig.level = alpha)

pwr.t.test(n = 20, d = 1, type = "two.sample",
           alternative="two.sided", sig.level = alpha)

pwr.t.test(power = 0.8, d = 1, sig.level=alpha)
```

### Question 4
Conclusion:
The probability to detect a difference of one standard deviation with 10 animals per group is 0.56;
with 20 animal, the probability is 0.87
To detect such difference with a probability of 80%, we need 17 animals per group (ceil(16.7) = 17).


## Multiple comparisons

  1. Redo the three tests of the previous section (questions 1-3) for now comparing the means of all the quantitative variables. Store the results (the p-values) in a data frame with one variable per row and four columns (name of the variable + three p-values). 
```{r}
do_all_comparisons <- function(data, alpha) {
  map(colnames(MON810_clean)[5:51], 
    function(col) {
    # question 1 
    males <- filter(data, period == 2 & sex == 'M') %>% pull(col)
    females <- filter(data, period == 2 & sex == 'F') %>% pull(col)
    t1 <- t.test(males, females, conf.level = 1-alpha, var.equal = FALSE)
    # question 2 
    period1 <- filter(data, period == 1 & sex == 'M') %>% pull(col)
    period2 <- filter(data, period == 2 & sex == 'M') %>% pull(col)
    t2 <- t.test(period1, period2, conf.level = 1-alpha, var.equal = FALSE)
    # question 3
    control <- filter (data, period == 2 & sex == 'M' & regimen == 'control') %>% pull(col)
    g_MON810 <- filter (data, period == 2 & sex == 'M' & regimen == 'MON810') %>% pull(col)
    t3 <- t.test(control, g_MON810, conf.level = 1-alpha, var.equal = FALSE)
    data.frame(
      Variable    = col,
      p.value_q1    = t1$p.value,
      p.value_q2    = t2$p.value,
      p.value_q3    = t3$p.value
    )
  }) %>% bind_rows()
}

all_comparisons <- do_all_comparisons(MON810_clean, alpha = 0.05)
all_comparisons %>% rmarkdown::paged_table()
```
   
   2. For each of the three tests, adjust the p-values using the Bonferroni  and the Benjamini-Hochberg corrections. How can we interpret these results?
```{r}
m <- nrow(all_comparisons)
all_comparisons$p.value_bonferonni_q1 <- pmin(1, all_comparisons$p.value_q1 * m)
all_comparisons$p.value_bonferonni_q2 <- pmin(1, all_comparisons$p.value_q2 * m)
all_comparisons$p.value_bonferonni_q3 <- pmin(1, all_comparisons$p.value_q3 * m)
all_comparisons$p.bh_q1 <- p.adjust(all_comparisons$p.value_q1, method = "BH")
all_comparisons$p.bh_q2 <- p.adjust(all_comparisons$p.value_q2, method = "BH")
all_comparisons$p.bh_q3 <- p.adjust(all_comparisons$p.value_q3, method = "BH")
all_comparisons %>% rmarkdown::paged_table()
```

### Question 3
Conclusion:
The p-values in question 1 and question 2 become bigger under Bonferroni corrections, as well as under Benjamini-Hochberg correction. 
And p-values in question 3 after correction are bounded by 1, or close to 1. That is to say, we are almost sure that there is a difference of all biochemical parameters between MON810 group and control group, for period 2, for males. 
  
# Linear models: quarterly sales volumes

## Fitting a linear model

The file `sales1.csv` consists of quarterly sales volumes (in % and indexed to the time 0) of a product.

```{r}
sales1 <- readr::read_csv("sales1.csv", show_col_types = FALSE)
rmarkdown::paged_table(sales1)
```

  1. Plot the data and comment.
```{r}
sales1_plot <- ggplot(data = sales1, aes(x = time, y = y)) +
  geom_point() + 
  xlab("Time") +  
  ylab("Quarterly sales volume") + 
  theme_bw()
sales1_plot
```

### Question 1
It seems that the sales volume increases with the time. We can try a linear model to fit it.

  2. Fit a polynomial model to this data (justify the choice of the degree). What do the residuals suggest? 
```{r}
sales1_plot + geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour="#339900")
model1 <- lm(y~time, data=sales1)
par(mfrow=c(2,2))
plot(model1)
summary(model1)
```

### Question 2
We choose polynomial model with degree 1, because the plot shows the linear relation between time and sales volume. 
Those 4 figures show that the residuals are centered gaussian, have homosedasctic variance, which corresponds well with the assumptions of gaussian linear regression. 
But the Residual standard error in summary shows that the residuals are relatively big. We can try to fit a better model.


  3. Try to improve the model by adding a periodic component. Write your final model as a mathematical equation.

::: {.callout-info} 

### Hints
  - $\cos(2\pi t/T)$ and $\sin(2\pi t/T)$ are periodic functions of period $T$.
  - One can easily deduce the period $T$ from the context and looking at the data.
:::

### Question 3
The improved model is:
$$ y = k*time + A*(cos(2*pi*time/T + phi)) $$

  4. Plot on a same graph the observed sales together with the predicted sales given by your final model. What do you think about this model? What about the residuals?
```{r}
# nlm model
nlm1 <- nls(y ~ k*time + b + A*cos(time*2*pi/T + phi), sales1, 
            start = c(k=0.3, b=100, A=5, T=10, phi=1))

pred_time <- data.frame(time = seq(0, 100, by=1))
pred_sales <- predict(nlm1, newdata=pred_time)
prediction1 <- pred_time %>% mutate(sales=pred_sales)

sales1_plot +
  geom_line(data = prediction1, aes(x = time, y = sales, colour="red"))
summary(nlm1)
```
### Question 4
This modified model represents well the periodic property of sales volume. 
Regarding the residuals, they are in average smaller compared to the previous linear model (1.05 compared to 2.2).


  5. We want the predicted sales volume to be equal to 100 at time 0. Modify your final model in order to take this constraint into account.
```{r}
nlm2 <- nls(y ~ k*time + A*(cos(time*2*pi/T + phi) - cos(phi)) + 100, sales1, 
            start = c(k=0.3, A=5, T=11, phi=5))

pred_time <- data.frame(time = seq(0, 100, by=1))
pred_sales2 <- predict(nlm2, newdata=pred_time)
prediction2 <- pred_time %>% mutate(sales=pred_sales2)

sales1_plot +
  geom_line(data = prediction2, aes(x = time, y = sales, colour="red"))
summary(nlm2)
```
### Question 5
The final model is:
$$y = k*time + A*(cos(2*pi*time/T + phi) - cos(phi)) + 100$$

## Fitting a linear mixed effects model

The file `sales30.csv` now consists of quarterly sales volumes (still in % and indexed to the time 0) of 30 different products.

```{r}
sales30 <- readr::read_csv("sales30.csv", show_col_types = FALSE) %>% mutate(id = as.factor(id))
rmarkdown::paged_table(sales30)
```

  1. Plot the data and comment
```{r}
sales30_plot <- ggplot(data = sales30, aes(x = time, y = y, color=id)) +
  geom_line() + facet_wrap(~id) +
  xlab("Time") +  
  ylab("Quarterly sales volume") + 
  theme_bw()
sales30_plot
```
### Question 1
Comment:
We find that for most products, the trend of sales volumes over time is similar to the previous case. 
But there are also several products whose sales volume is almost constant with a little fluctuations. For instance, products with id=12,13,14.

 
  2. Adjust the model used previously for fitting the first series to this data and comment the results.
```{r}
nlm30_1 <- nls(y ~ k*time + b + A*cos(time*2*pi/T + phi), sales30, 
            start = c(k=0.2, b=100, A=5, T=11, phi=1))
summary(nlm30_1)
```
```{r}
pred_time <- data.frame(time = seq(0, 100, by=1))
pred_sales30_1 <- predict(nlm30_1, newdata=pred_time)
prediction30_1 <- pred_time %>% mutate(sales=pred_sales30_1)

ggplot(data = sales30, aes(x = time, y = y)) +
  geom_line(size=1) + facet_wrap(~id) +
  theme_bw() +
  geom_line(data = prediction30_1, aes(x = time, y = sales), colour='#339900')
```

### Question 2
The green line is the fitted model. In some cases, it can reflect the general trend. But in some extreme cases, for instance, id=12,13,14, there is a big gap between the true value and predicted value. 
Therefore, we need to consider the random effect for each product.

  3. Fit a mixed effect model to this data (discuss the choice of fixed and random effects). Write your final model as a mathematical equation.
```{r}
saemix_data <- saemixData(name.data       = sales30,
                          name.group      = "id",
                          name.predictors = "time",
                          name.response   = "y")

model1_nlme <- function(psi,id,x) {
  t   <- x[,1]
  k  <- psi[id,1]
  A   <- psi[id,2]
  b  <- psi[id,3]
  T <-psi[id,4]
  phi <- psi[id,5]
  fpred <- k*t + b + A*cos(t*2*pi/T + phi)
  fpred
}

saemix_model1 <- saemixModel(model = model1_nlme,
                            psi0  = c(k=0.3,A=2,b=100,T=11,phi=1))

saemix_options <- list(map=TRUE, fim=TRUE, ll.is=FALSE, displayProgress=FALSE, save=FALSE, seed=632544)
saemix_fit1    <- saemix(saemix_model1, saemix_data, saemix_options)
```

### Question 3
We have to consider the random effect for all parameters. Because the plot of different products is quite different. Even the intercept is not the same.

The mixed effect model is:
$$ y_{ij} = f(t_{ij}, \psi_{i}) + \epsilon_{ij} \\where\ \psi_{i} = (k_{i}, A_{i}, b_{i}, T_{i}, phi_{i}) \\ f(t,\psi) = k*t + b + A*cos(t*2*pi/T + phi) \\for\ i = 1...N, j = 1...n_{i}$$
N is the number of products. n_i is the number of measurements for product i.


  4. Plot the data with the predicted sales given by the chosen model.
```{r}
saemix.plot.fits(saemix_fit1)
```

  5. Modify the previous model to take into account the previous constraint (predicted sales volume are all equal to 100 at time 0). 
```{r}
model2_nlme <- function(psi,id,x) {
  t   <- x[,1]
  k  <- psi[id,1]
  A   <- psi[id,2]
  T <-psi[id,3]
  phi <- psi[id,4]
  fpred <- k*t + A*(cos(t*2*pi/T + phi) - cos(phi)) + 100
  fpred
}

saemix_model2 <- saemixModel(model = model2_nlme,
                            psi0  = c(k=0.3,A=2,T=11,phi=1))
saemix_fit2 <- saemix(saemix_model2, saemix_data, saemix_options)
```

### Question 5
The final effect model is:
$$ y_{ij} = f(t_{ij}, \psi_{i}) + \epsilon_{ij} \\where\ \psi_{i} = (k_{i}, A_{i}, b_{i}, T_{i}, phi_{i}) \\ f(t,\psi) = k*t + A*(cos(t*2*pi/T + phi) - cos(phi)) + 100 \\for\ i = 1...N, j = 1...n_{i}$$
